Large inversions in Lake Malawi cichlids are associated with habitat preference, lineage, and sex determination

Version 1

Abstract

Abstract

Chromosomal inversions are an important class of genetic variation that link multiple alleles together into a single inherited block that can have important effects on fitness. To study the role of large inversions in the massive evolutionary radiation of Lake Malawi cichlids, we used long-read technologies to identify four single and two tandem inversions that span half of each respective chromosome, and which together encompass over 10% of the genome. Each inversion is fixed in one of the two states within the seven major ecogroups, suggesting they played a role in the separation of the major lake lineages into specific lake habitats. One exception is within the benthic sub-radiation, where both inverted and non-inverted alleles continue to segregate within the group. The evolutionary histories of three of the six inversions suggest they transferred from the pelagic Diplotaxodon group into benthic ancestors at the time the benthic sub-radiation was seeded. The remaining three inversions are found in a subset of benthic species living in deep waters. We show that some of these inversions are used as XY sex-determination systems but are also likely limited to a subset of total lake species. Our work suggests that inversions have been under both sexual and natural selection in Lake Malawi cichlids and that they will be important to understanding how this adaptive radiation evolved.

Introduction

Large genomic inversions are a particularly interesting type of genetic variation in the evolutionary process1–4. These rearrangements of DNA sequence strongly suppress recombination between inverted and non-inverted alleles, enabling the capture and accumulation of small genetic variants that are linked together on each inversion haplotype5. These alternative haplotypes largely follow independent evolutionary paths within a species, accumulating new mutations that can affect phenotypes under natural and sexual selection. Structural rearrangements can be large, spanning multiple megabases of DNA containing hundreds of genes2,6. They have been shown to play a role in adaptive divergence within a variety of species7–12, in creating alternative reproductive strategies and mating types13–16, in evolution of sex chromosomes1,17, and in the formation of pre- and post-zygotic barriers between incipient species18,19.

When comparing the genomes of extant species, chromosomal inversions are often found as fixed differences, suggesting they might have an important role in speciation20–22. However, the forces responsible for establishing inversions remain obscure. Do they play a role in adaptation to new ecological niches exploited by incipient species? Do they carry genetic variants that create prezygotic barriers that drive speciation? Can they resolve sexual conflict, linking sex-specific alleles to sex determining loci to create sexually dimorphic species? Perhaps most likely, they play roles in many or all these processes23,24.

Few vertebrates offer as much potential for investigating evolution and speciation as the Cichlidae family of fish, one of the most species-rich and diverse families of vertebrates, with an estimated 2,000-3,000 species across the globe 25–27. Evolutionary radiations have occurred at least three times in the African Great Lakes (Lake Malawi, Lake Tanganyika, and Lake Victoria). In Lake Malawi, more than 800 species of cichlids evolved in the past 1.2 million years26,28–30. Speciation in cichlid fishes involved extremely high levels of phenotypic divergence, including changes in body shape, jaw structures, and feeding behaviors related to prey in their ecological niches25,31. Additionally, traits that create prezygotic barriers between species are also extremely diverse. Color patterns are quite dramatic and diverse among species and include both differences in color and species-specific patterns including bars, stripes, and blotches that are thought to reinforce barriers between species in sympatry32. In some species, mating is seasonal, with gatherings on ‘leks’ where males build bowers by scooping and spitting out sand over the course of a week or more33. Females choose mates based upon features of these bowers including shape and size34. Most of the Lake Malawi species can be induced to interbreed in the lab, providing the opportunity to use genetic approaches to study causality of associated changes. Despite the presence of these prezygotic barriers, substantial gene flow has occurred multiple times within the lake29,35 and hybridization has been proposed to seed radiations within the African Great Lakes27,36,37.

Our current understanding of how the Lake Malawi radiation created >800 species is a set of three serial diversifications, starting from a single riverine-like ancestor lineage, that created seven major ecogroups separated primarily by habitat29. A pelagic lineage separated first, further diversifying into a deep-water ecogroup (Diplotaxodon) and a mid-water ecogroup (Rhamphochromis). A muddy/sandy benthic lineage evolved from the ancestral riverine lineage next, which split into three ecogroups (collectively referred to as benthics) living either in the water column (utaka), over shallow sandy/muddy shores (shallow benthics), or in deep-water habitats (deep benthics). Finally, an ecogroup living over rocky habitats evolved (mbuna). The riverine generalist Astatotilapia calliptera (AC) living in the border regions of the lake and surrounding rivers, is the seventh ecogroup, and is thought to represent the ancestral lineage that seeded the three primary Lake Malawi lineages. Subsequent radiations within these ecogroups based on trophic specializations and sexual selection further diversified the species flock29,31.

Here, we investigate the role large inversions played in the Lake Malawi radiation using new large molecule technologies. We identify four single large inversions and two double inversions, ranging from 9.9Mb to 20.6Mb in size. A fifth single inversion, composed of just one of the two tandem inversions on chromosome 20, was also identified, indicating a serial set of rearrangements created this structural variant. These inversions primarily segregate by ecogroup, with no inversions found in the mbuna and AC samples, one inversion fixed in Rhamphochromis, three inversions fixed in Diplotaxodon, and all six segregating within the benthic sub-radiation. The evolutionary histories of these inversions are inconsistent with the species phylogeny and suggest that the three Diplotaxodon inversions spread into the benthics via hybridization around the time the benthic lineage diverged from the ancestral riverine species. The three additional inversions are found primarily in benthic species living in deep water habitats. We provide evidence that three of the inversions are involved in sex determination in a subset of benthic species. Our work provides a framework for understanding the role of inversions in the adaptive radiation of Lake Malawi cichlids, and suggests multiple roles in the establishment of ecogroups, adaptation to deep water habitats, and in controlling traits under sexual selection.

Results

Improvement of the Metriaclima zebra reference

The mbuna, Metriaclima zebra, was the first species sequenced as a reference genome for Lake Malawi cichlids38. The current version of its genome, called M_zebra_UMD2a, has been created by assembling PacBio reads into contigs which were anchored to 22 chromosomes using recombinant maps generated using crosses between different Lake Malawi species39. While the overall quality of this reference is high compared to other non-model organisms, a total of 20.5% of the DNA was not anchored to a chromosome. Additionally, the orientation of contigs can be ambiguous given how contigs were anchored to linkage groups using recombinant maps. These imperfections could complicate inversion discovery, so we first sought to improve the reference sequence.

The Bionano Saphyr system generates large (50 kb - 1 Mb) DNA molecules and labels a specific motif (CTTAAG) with a fluorescent probe40. While these molecules do not provide sequencing information, the distance between observed fluorescent loci can be used to assemble multiple molecules into large maps, with lengths spanning up to an entire chromosome. Comparison of these maps with an in-silico digest of the reference sequence identifies large structural changes in the observed sample or errors in the reference genome. We processed DNA collected from the blood of three Metriaclima zebra individuals using the Bionano pipeline and assembled DNA molecules into optical maps. We identified many discrepancies between the observed optical maps and the M_zebra_UMD2a assembly (Figure 1). These differences fall into two primary categories: 1) contigs that were incorrectly ordered or oriented and 2) unplaced contigs that Bionano anchored to specific chromosomes. The first category is consistent with errors in the assembly caused by ambiguity in placing contigs into a linkage group using recombination maps. Our results indicate that 178 of the contigs are incorrectly oriented and 133 of the unanchored contigs could be placed onto specific chromosomes.

Generation of a new high quality reference genome for Metriaclima zebra.

Example alignment of Bionano maps created from blood isolated from a single Metriaclima zebra male to the M_zebra_UMD2a reference (top) or the newly generated M_zebra_GT3a reference (bottom). The M_zebra_GT3a reference was created using a combination of PacBio HiFi reads and Bionano molecules. A summary of the improvements in the new genome is found in Table S1.

To further improve the reference, we used PacBio HiFi reads generated from an additional male Metriaclima zebra individual to create a new genome assembly, taking advantage of the low overall error rate (< 0.1%) of HiFi reads compared to previous generation PacBio technologies. Contigs generated from these data were combined with Bionano maps to generate a new hybrid genome assembly, which we refer to as the M_zebra_GT3a assembly. The overall quality of the assembly was excellent, with 933Mb of 962 Mb of DNA assigned to the 22 linkage groups, (Table S1) reducing the amount of unplaced contigs from 196 Mb to 28 Mb. The N50 length was 32.542Mb, with many of the scaffolds containing entire chromosomes.

Identification of six large inversions from 11 species of Lake Malawi cichlids

Armed with the improved reference, we used the Bionano system to test eight different species of Lake Malawi cichlids that were currently used for experimental studies in the Streelman or McGrath laboratories (Table 1 and Table S2). We tested four mbuna species (Pseudotropheus demasoni, Cynotilapia zebroides, Labeotropheus fuelleborni, and Labeotropheus trewavasae), two shallow benthic species (Mchenga conophoros and Nyassachromis prostoma “Orange Cap”), a single deep benthic species (Aulonocara sp. ‘chitande type north’ Nkhata Bay), and a single utaka species (Copidachromis virginalis). To expand our coverage of the Lake Malawi radiation, we also obtained samples for a single species of the Diplotaxodon ecogroup, Diplotaxodon limnothrissa, and a single species of the Rhamphochromis ecogroup, Rhamphochromis longiceps. Finally, we also tested a single male Protomelas taeniolatus shallow benthic individual based on its position in principal component analysis of SNVs described below. In total, we were able to obtain samples from 6 of the 7 ecogroups, excluding Astatotilapia calliptera.

From these 31 individuals, we identified six new inversions ranging from 9.9 - 22.9 Mb in size (Figure 2): a single inversion on chromosomes 2, 10, 11, and 13, and a double inversion on chromosomes 11 and 20. Interestingly, regarding the double inversion on chromosome 20, we also identified four Rhamphochromis individuals that carried only the first inversion, lacking the 4.1 Mb long second part of this rearrangement (Figure 2). This indicates that the tandem inversions on 20 were formed by at least a two-part process, most parsimoniously with the left arm of the inversion (20a) forming first followed by the right inversion (20b) occurring in that genetic background.

Identification of six large single or double inversions segregating within Lake Malawi cichlids.

Representative alignments of inversions or double inversions identified from blood samples obtained from 30 individuals from eleven species. For each inversion, the top shows predicted motifs from an in-situ digest of the reference genome, the bottom shows motifs identified using Bionano molecules obtained from an individual of the species indicated on the left, and the grey lines indicate matching motifs based upon predicted and observed distances. Single inversions were identified on chromosomes 2, 9, 10, 13, and 20. Tandem double inversions were identified on chromosomes 11 and 20. The estimated length and percentage of the chromosome spanned by the inversion is shown on the right. The single inversion on 20 identified in Rhamphochromis longiceps has the same position as the first inversion of the double inversion on 20 identified from Mchenga conophoros. A list of all samples and their inversion genotype is in Table 1. Note that an error in the reference genome in chromosome 2 is indicated in the first panel.

The distribution within these species is summarized in Table 1. The most common structural rearrangements were the single inversion on chromosome 9, the double inversion on chromosome 11, and the double inversion on chromosome 20. These were found in a combination of benthic and Diplotaxodon individuals. The position of these three inversions were qualitatively consistent with regions of low recombination rate previously identified in an intercross between a mbuna and benthic individual, as expected, since inversions suppress recombination between inverted and non-inverted haplotypes22. The remaining three inversions (2, 10, and 13) were only found in the Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals. While for the majority of samples the inversion genotype was homozygous, for five samples optical maps supported both inverted and non-inverted haplotypes, indicating the inversions were heterozygous in those individuals (Figure S1).

To provide further support for these structural rearrangements, we generated de novo hybrid genome assemblies of three Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals, which carry all six of the identified inversions, and two additional Metriaclima zebra individuals using a combination of PacBio HiFi reads and Bionano maps. The overall quality of these assemblies is high, with large N50 values and a small number of individual contigs (Table S3). We performed whole genome alignments between the five new assemblies and the GT3a reference genome (Figure S2-S7). These alignments show that the new assemblies support the presence of all six inversions in the Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals. For most inversions, a single scaffold covered the entire inversion with breakpoints in locations consistent with the Bionano maps. The two new control assemblies were consistent with the M_zebra_GT3a reference and did not show any evidence of inversions in those regions. In other regions of the genome, however, there were differences in alignment of the M_zebra_GT3a assembly and the Metriaclima zebra genomes, most prominently in the left arm of chromosome 2 (Figure 2) and in the highly repetitive chromosome 3, which likely indicate some errors in the M_zebra_GT3a assembly.

We refined the location of the breakpoints for each inversion using the genomic alignments. However, the presence of a large amount of repetitive DNA in regions near the breakpoints prevented us from defining the inversion boundaries at single base pair resolution.

We also took advantage of a genome assembly for a Rhamphochromis sp ‘chilingali’ individual (fRhaChi2.1) created by the Darwin Tree of Life Project produced using PacBio data and Arima2 Hi-C data (Accession PRJEB72870). The genome alignment between the fRhaChi2.1 assembly and the GT3a assembly confirmed a single inversion on chromosome 20, consistent with the Bionano data (Figure S2-S7).

For each inversion, we determined the ancestral haplotype using two species as outgroups, Oreochromis niloticus and Pundamilia nyererei. Oreochromis niloticus, or Nile tilapia, is an important food fish that’s estimated to have diverged from Lake Malawi cichlids 14.1 to 30 million years ago while Pundamilia nyererei is a more closely related haplochromine found in Lake Victoria38,41. For this analysis we used a published whole genome assembly for Oreochromis niloticus (UMD_NMBU)39 and generated a hybrid genome assembly for Pundamilia nyererei using a previously published whole genome assembly38 combined with Bionano maps we produced from a single individual (Figure S8). For all six inversions, the reference haplotype found in the GT3a assembly represents the ancestral state.

Inference of the segregation of these inversions in Lake Malawi using short-read sequencing

The distribution of these inversions within the lake can provide important information to their function. Genotyping inversions using short read sequencing is possible because each haplotype is on an independent evolutionary trajectory due to the suppression of recombination. Population genetics approaches take advantage of the fact that inversions create patterns of inheritance between SNVs that are readily detected by approaches such as Principal Component Analysis (PCA)42.

We analyzed short read sequencing data from the 31 samples with Bionano data along with 297 wild individuals published by the Durbin lab29, 15 wild individuals published by the Streelman lab 30,43, and 22 wild individuals sequenced for this paper, aligning reads and calling small variants against our new M_zebra_GT3a reference genome. PCA analysis was used to analyze the major axes of variation across the whole genome and within the six inversions. We separately analyzed the left and right inversions on 11 and 20 (the two inversions on 11 were identical in their distribution). For the whole genome, the distribution of ecogroups within the first two principal components matched the previously reported results of Malinsky et al29. Individuals from four of the ecogroups - Rhamphochromis, Diplotaxodon, Astatotilapia calliptera, and mbuna - formed four distinct clusters (Figure 3A and Figure S9). A fifth, more diffuse cluster, composed of shallow benthics, deep benthics, and utakas, was also present. The broader distribution of the benthic ecogroups in PC space is consistent with both their recent separation as well as large amounts of gene flow thought to have occurred between these different species29,35.

Distribution of inversions within the Lake Malawi ecogroups.

Principal component analysis was used to analyze SNVs identified using whole genome sequencing from 365 samples to genotype the inversion haplotype for all six inversions. (A) We illustrate the approach for genotyping the chromosome 10 inversions. The PCA plot for the entire genome is shown on the left, with each sample colored by its ecogroup. Individuals with Bionano data are shown as grey Xs. The deep benthic/shallow benthic and utaka individuals that cluster together for the whole genome analysis split into multiple clusters when analyzed using the SNVs that fall within the chromosome 10 inversion, and these clusters were assigned to inversion genotypes using samples that were genotyped using Bionano data. To make comparisons of the whole genome PCA and chromosome 10 PCA plots easier, we reversed the y-axis on the right panel. The genotypes of each sample can be found in Supplemental Table 2. Interactive PCA plots for each inversion are included in Figures S9-S17. (B) The derived allele frequency was calculated for each inversion within the seven ecogroups. The number of species that were genotyped and the estimated number of species that live in Lake Malawi are listed in the whole-genome phylogeny (left).

The PCA plots using the SNVs within the six inversions often violated the pattern of the whole genome PCA (Figure 3A and Figure S10-S17). Using samples that were sequenced with both Illumina and Bionano technologies, we assigned each cluster to specific inversion genotypes (Table S2). We did not have Bionano data for two clusters (one on 9 and one on 11 – see Figures S11, S13, and S14), so we supplemented our dataset by adding Bionano data from an additional shallow benthic individual (Protomelas taeniolatus) that fell into both clusters. The PCA clusters varied in their complexities for each inversion depending largely on their distribution within the benthic ecogroup (more on this below) and included both homozygous genotypes and heterozygous genotypes. We summarize the overall distribution of the inversions within each of the ecogroups in Figure 3B.

All specimens of the mbuna and AC ecogroup were fixed for the homozygous non-inverted alleles for all six inversions. In Rhamphochromis, all the inversions were homozygous non-inverted except for the left arm of 20 which was fixed for the single inversion state. The Diplotaxodon individuals all carried the non-inverted alleles for the inversions on 2, 10 and 13 and were fixed for the inversions on 9, 11, and 20.

The distribution of these inversions was the most complicated within the benthics. The inversions on 2, 10, and 13, which were identified in Aulonocara sp. ‘chitande type north’ Nkhata Bay, were mostly found together in other deep benthics, including the Alticorpus (geoffreyi, macrocleithrum, peterdaviesi), Aulonocara (blue chilumba, gold and minutus), and Lethrinops (gossei, longimanus ‘redhead’, and sp. olivera) genera. The inversions on 2 and 13 were also found in the shallow benthic species Placidochromis longimanus. However, not all the deep benthics carried the three inversions, including five Aulonocara species (baenschi, getrudae, steveni, stuartgranti, and stuartgranti Maisoni) and a single Lethrinops species (longipinnis). The distribution of these three inversions in benthic species known to inhabit the deepest habitats are suggestive for a role in depth adaptation.

The double inversion on 20 was homozygous inverted in all benthic individuals. For the single inversion on 9, many benthics were homozygous for the inverted haplotype. However, an additional PC cluster was present, composed of 64 shallow benthic and utaka individuals heterozygous for the non-inverted haplotype (Figure S1). Interestingly, no benthic individuals were identified as homozygous for the non-inverted state.

Finally, the distribution of the chromosome 11 double inversion was also determined. Individuals of all three genotypes (homozygous non-inverted, homozygous inverted, and heterozygous) were observed (Figure 3a).

Evolutionary history of 9,11, and 20 suggest introgression from Diplotaxodon into benthic ancestors

The distribution of these inversions conflicts with the species tree, suggesting that these inversions could have spread via hybridization. To determine their evolutionary history, we built phylogenies for the whole genome as well as for each inversion using SNVs that fell within each of the six inversions using representative species from each ecogroup (73 total – Table S2) as well as two previously sequenced Pundamilia nyererei individuals as an outgroup (Figure 4A and Figures S18 – S25)44,45. To avoid issues with haplotype phasing, we focused on individuals that were homozygous for one of the two inverted genotypes.

Evolutionary history of large inversions violates species tree.

(A) Maximum likelihood trees built from either whole genome SNVs (top) or within the 9, 11, and 20 inversions. Between 2-25 representative samples were selected for each ecogroup (see Table S2) and individuals that were heterozygous were excluded from this analysis. Clades were collapsed by ecogroup for visualization purposes. Full phylogenies are presented in Figures S18-S25. The whole genome tree indicates Diplotaxodon and Rhamphochromis species split first and formed their own clade. For each of the displayed inversions, the first split was between samples carrying the inverted vs. normal genotypes. Additionally, the Diplotaxodon individuals now formed a clade with the benthics that carry the inversion. (B) Density plots of dXY values comparing benthic species and species of the indicated ecogroup. For the displayed inversions, the Diplotaxodon species were much closer in evolutionary distance when compared to the whole genome plot.

While the whole genome phylogeny matched the expected topology based upon our understanding of the Lake Malawi species tree, the topologies of each of the inversions showed key differences between the separation of ecogroups. The benthic group often split into separate clades, correlated with their inversion genotype. Prominently, we found the three inversions on 9, 11, and 20 showed key differences in the relationship between the Diplotaxodon and benthic ecogroups. While Diplotaxodon is a sister group to Rhamphochromis at the species level, for these three inversions, the Diplotaxodon individuals formed a clade with the benthics that also carried the inversion. Additionally, genetic distance between Diplotaxodon and benthics carrying the inversion was also much smaller than the rest of the genome (Figure 4B). Both the phylogenies and genetic distances between ecogroups are consistent with the introgression of the 9, 11, and 20 inversions from the Diplotaxodon to benthic ancestors at the time of the benthic radiation.

A role for large inversions in sex determination within the benthic ecogroup

The evolution of sex determination is often associated with the presence of inversions to repress recombination between the sex chromosomes46,47. We tested whether these inversions could play a role in sex determination in benthic species, as we identified individuals that were heterozygous for some of these inversions (Table 1). We started with the inversion on 10, as our Bionano data identified three male Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals that were heterozygous and a single female that was homozygous for the inverted haplotype. To test whether this inversion acted as an XY system in this species, we tested two separate broods (Brood 1; 2 parents, 24 offspring and Brood 2: 2 parents, 24 offspring) currently growing in the laboratory, sequencing 12 males and 12 females from each brood with Illumina paired-end sequencing. We used PCA to genotype each of the 52 animals (4 parents and 48 offspring) for the chromosome 10 inversion genotype. The association in the offspring between the inversion 10 genotype and sex was almost perfect: 24 females and 1 male were homozygous for the inversion, while 23 males were heterozygous for the inversion (Figure 5B).

A role for the chromosome 10 inversion in sex determination

(A) Schematic of the two inversion states (normal and inverted) and their assignment as X or Y chromosomes. (B) 24 o‘spring (12 male and 12 female) from two separate male/female pairs were genotyped for the inversion. For 47 of 48 individuals, the XY animals were male and the XX animals were female. One male was XX. (C) FST analysis comparing 24 males and 24 females identified chromosome 10 as the primary genetic region that segregated with sex. (D) Comparison of heterozygosity leve2ls9o3f 24 male and 24 female animals also identified chromosome 10 as an outlier.

We also used binned Weir Cockerham Fst analysis, grouping male and female individuals to scan the genome for regions that were associated with sex (Figure 5C). The inversion on 10 was the region of the genome that was in strongest linkage disequilibrium with sex. Males also had a much higher heterozygosity in this region than females (Figure 5D). Altogether, these results indicate that the inversion on 10 acts as an XY sex determiner in the Aulonocara sp. ‘chitande type north’ Nkhata Bay species.

We also tested eight laboratory Nyassachromis prostoma “Orange Cap” individuals, as our Bionano data identified a male that was heterozygous for the inversion on 11 and a female that was homozygous for the inversion. We performed short-read sequencing on three males and five females growing in the lab, genotyping the inversion using PCA analysis. All three males were heterozygous for the inversion while the five females were homozygous, a significant association between genotype and sex (Fisher exact test p-value = 0.0179). This suggests a separate XY system segregates in Nyassachromis prostoma “Orange Cap” using the inversion on chromosome 11.

Finally, for the inversion on nine, there was a cluster of heterozygous individuals that included 7 individuals that we had sex information for. All 7 of these individuals were male (Table S2). While this association is not significant, it is also suggestive that the inversion on 9 could play a role in sex determination as well.

Discussion

Here, we characterize the presence of large inversions found in the Lake Malawi cichlid flock, identifying four single inversions and two double inversions, providing a framework to address the role structural variants played in this lake’s adaptive radiation (Figure 6). Inversions can have multiple roles in the evolutionary process: 1) they can capture beneficial alleles responsible for adaptation to local habitats, 2) they can capture alleles that create pre- and post-zygotic barriers that prevent gene flow during speciation, and 3) they can capture alleles that influence sex-specific phenotypes that are linked to sex-determination systems1,3,4. Future work is now possible to determine how these inversions contribute to the evolution of Lake Malawi cichlids through these mechanisms.

Model for inversions in Lake Malawi.

Model for the origin spread of inversions in Lake Malawi overlayed on the species tree proposed in Malinsky et al. Key events in the proposed inversions’ histories are indicated by numbers overlayed on the phylogeny.

The overall size of these inversions is large, together capturing >10% of the genome (∼112 Mb). Apart from the benthics, the inversions are fixed in each ecogroup in the inverted or non-inverted state. Such a distribution would be expected in the local adaptation hypothesis if they were involved in the adaptation of these ecogroups to their respective lake habitats. In this scenario, the inversions would be under selection prior to speciation, and would play roles in: 1) the split of the pelagic group from the ancestral riverine lineage (the left arm of the inversion on 20), 2) the diversification of the Rhamphochromis and Diplotaxodon groups (the inversions on 9, 11, and right arm of 20), and 3) the adaptation of benthics to deeper water habitats (the inversions on 2, 10, and 13). The role of inversions in local adaptation between spatially separated populations has been well characterized6,8. If two populations are undergoing adaptive divergence in the presence of migration and gene flow, theory indicates that inversions will be selected for when they capture and link adaptive alleles together48. Given gene flow was likely common during the formation of the ecogroups49, we expect inversions that captured adaptive alleles that increase fitness in a habitat-specific context to feature prominently in ecogroup inception. This hypothesis predicts that adaptive alleles for each ecogroup are enriched within the inversions, which can be tested in Lake Malawi cichlids, as different species can interbreed to test whether the inversions are associated with differences in phenotypes.

While the role of the inversions in most of the ecogroups could potentially be explained by local adaptation, their role in the benthic sub-radiation appears to be more complicated. Inverted and non-inverted haplotypes for five of the six inversions are segregating within this group. Additionally, the presence of the 9, 11, and 20 inversions within benthics violates the whole species phylogeny and is consistent with its spread from the Diplotaxodon to a very early benthic ancestor via introgression between ancestors of these two groups. Hybridization has been proposed to fuel adaptive radiations of cichlids in all three great African lakes (Malawi, Tanganyika, and Victoria) and our work suggests that an ancient hybridization event between early benthic and Diplotaxodon ancestors occurred prior to the benthic radiation resulting in the 3 inversions (as well as other regions of the genome) spreading into what ultimately became the benthic ecogroup26,27,36. It will be interesting to study the role of these three inversions in benthic species. Diplotaxodon are pelagic, deep-water dwellers that feed on plankton or small fish while benthics live near the shore over sandy and muddy habitats29. It is not obvious how the traits of the Diplotaxodon would benefit benthic animals that live in such dissimilar environments. However, the initial habitats that the benthic ancestors utilized is not known and potentially their initial habitat was in deeper waters than the AC ancestors. Whether or not these inversions were involved in the benthic radiations, they are not strictly required for large radiations as the mbuna group, with an estimated >300 species, carries zero large inversions29. This could reflect the relative importance of allopatric vs. sympatric speciation in the two radiations, as the rocky habitats the mbuna inhabit are often spatially separated throughout the lake.

The inversions we identified also play a role in sex determination in the benthics, with at least three of the inversions (9, 10, and 11) acting as XY systems to control sex determination. Despite the importance of the maintenance of distinct, reproductively compatible sexes, sex determination systems can be evolutionarily labile, and Lake Malawi cichlids show exceptional variability, with at least 5 different genetic sex determination systems described to date50,51. While we implicated multiple inversions in sex determination, our work also underscores how dynamic sex determination is within the benthics. While many benthic species carry the different inversions, their genotypic distribution in most species is inconsistent with a role for sex determination. Effectively, the genetic determination of sex determination largely must be determined in a species by species way, and individual species may be segregating multiple sex determination systems. Additionally, these inversions could have played a role in sex determination in the ancestors of pelagic ecogroups, which could have contributed to their spread before sex chromosome turnover ended balancing selection on the inverted/non-inverted haplotypes and subsequent fixation of these inversions. Identification of the causal alleles for sex determination on these inversions will allow us to resolve this question.

The presence of non-inverted haplotypes in the benthics could have been retained from the time the lineage initially formed, with both haplotypes under balancing selection if they were used for sex determination. A recent preprint, however, which identified 5 of the 6 inversions presented here via short-read sequencing of 1375 individuals sampled from Lake Malawi, presented an alternative model52. They proposed the non-inverted haplotypes of the 9 and 10 inversions were introgressed back into the benthics from riverine species within and outside of the lake. Similarly, the non-inverted haplotypes on chromosomes 2, 10, and 13 in the deep benthics were introgressed from shallow benthics. In this model, it is unclear what evolutionary forces drove these back introgressions of the non-inverted alleles and when they developed a role in sex determination.

The selective forces on sex chromosomes are strong, including balancing selection for sex ratio and sexually antagonistic selection on sexually dimorphic phenotypes. Lake Malawi cichlids show amazing sex differences, characterized by diversification in numerous phenotypes, including pigmentation, behavior, and morphology thought to contribute to species barriers. The use of inversions in sex-determination allows for the accumulation of alleles in linkage with the sex-determiner gene which will have sex-specific effects in the male individuals that carry the Y chromosomes. Interestingly in all three cases, the non-inverted haplotype acts as the Y, suggesting that different haplotypes could play independent roles in evolution. For example, the inversion X haplotype on 10 could carry alleles responsible for adaptation to deep water habits, which would be carried by both sexes, while the non-inverted Y haplotype carries alleles under sexual selection.

This report provides an initial framework for understanding the role of inversions in the adaptive radiation of Lake Malawi cichlids. Future work identifying and characterizing causal alleles captured within these inversions will enable us to discern the specific mechanisms by which they contribute to speciation using this remarkable evolutionary system as a model for the evolution of all animals.

eLife Assessment

Using several hundreds of samples and cutting-edge genomic methods, including BioNano, PacBio HiFi, and advanced bioinformatic pipelines, the authors identify six large chromosomal inversions segregating in over 100 species of Lake Malawi cichlids. This important study provides compelling evidence for the presence of these six inversions, their differential distribution among populations, and the association of chromosome 10 inversion with a sex-determination locus. This work also provides a starting point for further investigating the role of these inversions with respect to local adaptation, speciation, sex determination, hybridization, and incomplete lineage sorting in cichlids, which represent ~5% of the extant vertebrate species and are one of the most prominent examples of adaptive radiations.

Reviewer #1 (Public review):

Summary:

Using high-quality genomic data (long-reads, optical maps, short-reads) and advanced bioinformatic analysis, the authors aimed to document chromosomal rearrangements across a recent radiation (Lake Malawi Cichlids). Working on 11 species, they achieved a high-resolution inversion detection and then investigated how inversions are distributed within populations (using a complementary dataset of short-reads), associated with sex, and shared or fixed among lineages. The history and ancestry of the inversions is also explored.

On one hand, I am very enthusiastic about the global finding (many inversions well-characterized in a highly diverse group!) and impressed by the amount of work put into this study. On the other hand, I have struggled so much to read the manuscript that I am unsure about how much the data supports some claims. I'm afraid most readers may feel the same and really need a deep reorganisation of the text, figures, and tables. I reckon this is difficult given the complexity brought by different inversions/different species/different datasets but it is highly needed to make this study accessible.

The methods of comparing optical maps, and looking at inversions at macro-evolutionary scales can be useful for the community. For cichlids, it is a first assessment that will allow further tests about the role of inversions in speciation and ecological specialisation. However, the current version of the manuscript is hardly accessible to non-specialists and the methods are not fully reproducible.

Strengths:

(1) Evidence for the presence of inversion is well-supported by optical mapping (very nice analysis and figure!).

(2) The link between sex determination and inversion in chr 10 in one species is very clearly demonstrated by the proportion in each sex and additional crosses. This section is also the easiest to read in the manuscript and I recommend trying to rewrite other result sections in the same way.

(3) A new high-quality reference genome is provided for Metriaclima zebra (and possibly other assemblies? - unclear).

(4) The sample size is great (31 individuals with optical maps if I understand well?).

(5) Ancestry at those inversions is explored with outgroups.

(6) Polymorphism for all inversions is quantified using a complementary dataset.

Weaknesses:

(1) Lack of clarity in the paper: As it currently reads, it is very hard to follow the different species, ecotypes, samples, inversions, etc. It would be useful to provide a phylogeny explicitly positioning the samples used for assembly and the habitat preference. Then the text would benefit from being organised either by variant or by subgroups rather than by successive steps of analysis.

(2) Lack of information for reproducibility: I couldn't find clearly the filters and parameters used for the different genomic analyses for example. This is just one example and I think the methods need to be re-worked to be reproducible. Including the codes inside the methods makes it hard to follow, so why not put the scripts in an indexed repository?

(3) Further confirmation of inversions and their breakpoints would be valuable. I don't understand why the long-reads (that were available and used for genome assembly) were not also used for SV detection and breakpoint refinement.

(4) Lack of statistical testing for the hypothesis of introgression: Although cichlids are known for high levels of hybridization, inversions can also remain balanced for a long time. what could allow us to differentiate introgression from incomplete lineage sorting?

(5) The sample size is unclear: possibly 31 for Bionano, 297 for short-reads, how many for long-reads or assemblies? How is this sample size split across species? This would deserve a table.

(6) Short read combines several datasets but batch effect is not tested.

(7) It is unclear how ancestry is determined because the synteny with outgroups is not shown.

(8) The level of polymorphism for the different inversions is difficult to interpret because it is unclear whether replicated are different species within an eco-group or different individuals from the same species. How could it be that homozygous references are so spread across the PCA? I guess the species-specific polymorphism is stronger than the ancestral order but in such a case, wouldn't it be worth re-doing the PCa on a subset?

Reviewer #2 (Public review):

Summary:

Chromosomal inversions have been predicted to play a role in adaptive evolution and speciation because of their ability to "lock" together adaptive alleles in genomic regions of low recombination. In this study, the authors use a combination of cutting-edge genomic methods, including BioNano and PacBio HiFi sequencing, to identify six large chromosomal inversions segregating in over 100 species of Lake Malawi cichlids, a classic example of adaptive radiation and rapid speciation. By examining the frequencies of these inversions present in species from six different linages, the authors show that there is an association between the presence of specific inversions with specific lineages/habitats. Using a combination of phylogenetic analyses and sequencing data, they demonstrate that three of the inversions have been introduced to one lineage via hybridization. Finally, genotyping of wild individuals as well as laboratory crosses suggests that three inversions are associated with XY sex determination systems in a subset of species. The data add to a growing number of systems in which inversions have been associated with adaptation to divergent environments. However, like most of the other recent studies in the field, this study does not go beyond describing the presence of the inversions to demonstrate that the inversions are under sexual or natural selection or that they contribute to adaptation or speciation in this system.

Strengths:

All analyses are very well done, and the conclusions about the presence of the six inversions in Lake Malawi cichlids, the frequencies of the inversions in different species, and the presence of three inversions in the benthic lineages due to hybridization are well-supported. Genotyping of 48 individuals resulting from laboratory crosses provides strong support that the chromosome 10 inversion is associated with a sex-determination locus.

Weaknesses:

The evidence supporting a role for the chromosome 11 inversion and the chromosome 9 inversion in sex determination is based on relatively few individuals and therefore remains suggestive. The authors are mostly cautious in their interpretations of the data. However, there are a few places where they state that the inversions are favored by selection, but they provide no evidence that this is the case and there is no consideration of alternative hypotheses (i.e. that the inversions might have been fixed via drift).

Reviewer #3 (Public review):

This is a very interesting paper bringing truly fascinating insight into the genomic processes underlying the famous adaptive radiation seen in cichlid fishes from Lake Malawi. The authors use structural and sequence information from species belonging to distinct ecotypic categories, representing subclades of the radiation, to document structural variation across the evolutionary tree, infer introgression of inversions among branches of the clade, and even suggest that certain rearrangements constitute new sex-determining loci. The insight is intriguing and is likely to make a substantial contribution to the field and to seed new hypotheses about the ecological processes and adaptive traits involved in this radiation.

I think the paper could be clarified in its prose, and that the discussion could be more informative regarding the putative roles of the inversions in adaptation to each ecotypic niche. Identifying key, large inversions shared in various ways across the different taxa is really a great step forward. However, the population genomics analysis requires further work to describe and decipher in a more systematic way the evolutionary forces at play and their consequences on the various inversions identified.

The model of evolution involving multiple inversions putatively linking together co-adapted "cassettes" could be better spelled out since it is not entirely clear how the existing theory on the recruitment of inversions in local adaptation (e.g. Kirkpatrick and Barton) operates on multiple unlinked inversions. How such loci correspond to distinct suites of integrated traits, or not, is not very easy to envision in the current state of the manuscript.

The role of one inversion in sex determination is apparent and truly intriguing. However, the implication of such locus on ecological adaptation is somewhat puzzling. Also, whether sex determination loci can flow across species via introgression seems quite important as a route to chromosomal sex determination, so this could be discussed further.

Author response:

The following is the authors’ response to the original reviews

We thank the reviewers for the careful review of our manuscript. Overall, they were positive about our use of cutting-edge methods to identify six inversions segregating in Lake Malawi. Their distribution in ~100 species of Lake Malawi species demonstrated that they were differentially segregating in different ecogroups/habitats and could potentially play a role in local adaptation, speciation, and sex determination. Reviewers were positive about our finding that the chromosome 10 inversion was associated with sex-determination in a deep benthic species and its potential role in regulating traits under sexual selection. They agree that this work is an important starting point in understanding the role of these inversions in the amazing phenotypic diversity found in the Lake Malawi cichlid flock.

There were two main criticisms that were made which we summarize:

(1) Lack of clarity. It was noted that the writing could be improved to make many technical points clearer. Additionally, certain discussion topics were not included that should be.

We will rewrite the text and add additional figures and tables to address the issues that were brought up in a point-by-point response. We will improve/include (1) the nomenclature to understand the inversions in different lineages, (2) improved descriptions for various genomic approaches, (3) a figure to document the samples and technologies used for each ecogroup, and 4) integration of LR sequences to identify inversion breakpoints to the finest resolution possible.

(2) We overstate the role that selection plays in the spread of these inversions and neglect other evolutionary processes that could be responsible for their spread.

We agree with the overarching point. We did not show that selection is involved in the spread of these inversions and other forces can be at play. Additionally, there were concerns with our model that the inversions introgressed from a Diplotaxodon ancestor into benthic ancestors and incomplete lineage sorting or balancing selection (via sex determination) could be at play. Overall, we agree with the reviewers with the following caveats. 1. Our analysis of the genetic distance between Diplotaxodons and benthic species in the inverted regions is more consistent with their spread through introgression versus incomplete lineage sorting or balancing selection. 2. Further the role of these inversions is likely different in different species. For example, the inversion of 10 and 11 play a role in sex determination in some species but not others and the potential pressures acting on the inverted and non-inverted haplotypes will be very different. These are very interesting and important questions booth for understanding the adaptive radiations in Lake Malawi and in general, and we are actively studying crosses to understand the role of these inversions in phenotypic variation between two species. We will modify the text to make all of these points clearer.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Using high-quality genomic data (long-reads, optical maps, short-reads) and advanced bioinformatic analysis, the authors aimed to document chromosomal rearrangements across a recent radiation (Lake Malawi Cichlids). Working on 11 species, they achieved a high-resolution inversion detection and then investigated how inversions are distributed within populations (using a complementary dataset of short-reads), associated with sex, and shared or fixed among lineages. The history and ancestry of the inversions is also explored.

On one hand, I am very enthusiastic about the global finding (many inversions well-characterized in a highly diverse group!) and impressed by the amount of work put into this study. On the other hand, I have struggled so much to read the manuscript that I am unsure about how much the data supports some claims. I'm afraid most readers may feel the same and really need a deep reorganisation of the text, figures, and tables. I reckon this is difficult given the complexity brought by different inversions/different species/different datasets but it is highly needed to make this study accessible.

The methods of comparing optical maps, and looking at inversions at macro-evolutionary scales can be useful for the community. For cichlids, it is a first assessment that will allow further tests about the role of inversions in speciation and ecological specialisation. However, the current version of the manuscript is hardly accessible to non-specialists and the methods are not fully reproducible.

Strengths:

(1) Evidence for the presence of inversion is well-supported by optical mapping (very nice analysis and figure!).

(2) The link between sex determination and inversion in chr 10 in one species is very clearly demonstrated by the proportion in each sex and additional crosses. This section is also the easiest to read in the manuscript and I recommend trying to rewrite other result sections in the same way.

(3) A new high-quality reference genome is provided for Metriaclima zebra (and possibly other assemblies? - unclear).

(4) The sample size is great (31 individuals with optical maps if I understand well?).

(5) Ancestry at those inversions is explored with outgroups.

(6) Polymorphism for all inversions is quantified using a complementary dataset.

Weaknesses:

(1) Lack of clarity in the paper: As it currently reads, it is very hard to follow the different species, ecotypes, samples, inversions, etc. It would be useful to provide a phylogeny explicitly positioning the samples used for assembly and the habitat preference. Then the text would benefit from being organised either by variant or by subgroups rather than by successive steps of analysis.

We have extensively rewritten the paper to improve the clarity. With respect to this point, we moved Figure 6 to Figure 1, which places the phylogeny of Lake Malawi cichlids at the beginning of the paper. We incorporated information about samples/technologies by ecogroup into this figure to help the reader gain an overview of the technologies involved. We added information about habitat for each ecogroup as well. While we considered a change to the text organization suggested here, we thought it was clearer to keep the original headings.

(2) Lack of information for reproducibility: I couldn't find clearly the filters and parameters used for the different genomic analyses for example. This is just one example and I think the methods need to be re-worked to be reproducible. Including the codes inside the methods makes it hard to follow, so why not put the scripts in an indexed repository?

We now provide a link to a github repository (https://github.com/ptmcgrat/CichlidSRSequencing/tree/Kumar_eLife) containing the scripts used for the major analysis in the paper. Because our data is behind a secure Dropbox account, readers will not be able to run the analysis, however, they can see the exact programs, filters, and parameters used for manuscript embedded within each script.

(3) Further confirmation of inversions and their breakpoints would be valuable. I don't understand why the long-reads (that were available and used for genome assembly) were not also used for SV detection and breakpoint refinement.

We did use long reads to confirm the presence of the inversions by creating five new genome assemblies from the PacBio HiFi reads: two additional Metriaclima zebra samples and three Aulonocara samples. Alignment of these five genomes to the MZ_GT3 reference is shown in Figures S2 – S7. These genome assemblies were also used to identify the breakpoints of the inversions. However, because of the extensive amount of repetitive DNA at the breakpoints (which is known to be important for the formation of large inversions), our ability to resolve the breakpoints was limited.

(4) Lack of statistical testing for the hypothesis of introgression: Although cichlids are known for high levels of hybridization, inversions can also remain balanced for a long time. what could allow us to differentiate introgression from incomplete lineage sorting?

The coalescent time between the inversions between Diplotaxodons and benthics should allow us to distinguish these two mechanisms. Our finding that the genetic distance, which is related to coalescent time, is closer within the inversions than the whole genome is supportive of introgression. However, we did not perform any simulations or statistical tests. We make it clearer in the text that incomplete lineage sorting remains a possible mechanism for the distribution of inversions within these ecogroups.

(5) The sample size is unclear: possibly 31 for Bionano, 297 for short-reads, how many for long-reads or assemblies? How is this sample size split across species? This would deserve a table.

We have included this information in the new Figure 1.

(6) Short read combines several datasets but batch effect is not tested.

We do not test for batch effect. However, we do note that all of the datasets were analyzed by the same pipeline starting from alignment so batch effects would be restricted to aspects of the reads themselves. Additionally, samples from the different data sets clustered as expected by lineage and inferred inversion, so for these purposes unlikely to have affected analysis.

(7) It is unclear how ancestry is determined because the synteny with outgroups is not shown.

Ancestry analysis was determined using the genome alignments of two outgroups from outside of Lake Malawi. This is shown in Figure S8.

(8) The level of polymorphism for the different inversions is difficult to interpret because it is unclear whether replicated are different species within an eco-group or different individuals from the same species. How could it be that homozygous references are so spread across the PCA? I guess the species-specific polymorphism is stronger than the ancestral order but in such a case, wouldn't it be worth re-doing the PCa on a subset?

The genomic PCA plots reflect the evolutionary histories that are observed in the whole genome phylogenies. Because the distribution of the inverted alleles violate the species tree, they form separate clusters on the PCA plots that can be used to genotype specific species. We have also performed this analysis on benthics (utaka/shallow benthics/deep benthics) and the distribution matches the expectation.

Reviewer #2 (Public review):

Summary:

Chromosomal inversions have been predicted to play a role in adaptive evolution and speciation because of their ability to "lock" together adaptive alleles in genomic regions of low recombination. In this study, the authors use a combination of cutting-edge genomic methods, including BioNano and PacBio HiFi sequencing, to identify six large chromosomal inversions segregating in over 100 species of Lake Malawi cichlids, a classic example of adaptive radiation and rapid speciation. By examining the frequencies of these inversions present in species from six different linages, the authors show that there is an association between the presence of specific inversions with specific lineages/habitats. Using a combination of phylogenetic analyses and sequencing data, they demonstrate that three of the inversions have been introduced to one lineage via hybridization. Finally, genotyping of wild individuals as well as laboratory crosses suggests that three inversions are associated with XY sex determination systems in a subset of species. The data add to a growing number of systems in which inversions have been associated with adaptation to divergent environments. However, like most of the other recent studies in the field, this study does not go beyond describing the presence of the inversions to demonstrate that the inversions are under sexual or natural selection or that they contribute to adaptation or speciation in this system.

Strengths:

All analyses are very well done, and the conclusions about the presence of the six inversions in Lake Malawi cichlids, the frequencies of the inversions in different species, and the presence of three inversions in the benthic lineages due to hybridization are well-supported. Genotyping of 48 individuals resulting from laboratory crosses provides strong support that the chromosome 10 inversion is associated with a sex-determination locus.

Weaknesses:

The evidence supporting a role for the chromosome 11 inversion and the chromosome 9 inversion in sex determination is based on relatively few individuals and therefore remains suggestive. The authors are mostly cautious in their interpretations of the data. However, there are a few places where they state that the inversions are favored by selection, but they provide no evidence that this is the case and there is no consideration of alternative hypotheses (i.e. that the inversions might have been fixed via drift).

We have removed mention of chromosome 9’s potential role in sex determination from the paper. While our analysis of sex association with chromosome 11 was limited compared to our analysis of chromosome 10, it was still statistically significant, and we believe it should be left in the paper. The role of 11 (and 9 and 10) in sex determination was also demonstrated using an independent dataset by Blumer et al (https://doi.org/10.1101/2024.07.28.605452)

We agree that we did not properly consider alternative hypothesis in the original submission and have rewritten the Discussion substantially to consider various alternative hypothesis.

Reviewer #3 (Public review):

This is a very interesting paper bringing truly fascinating insight into the genomic processes underlying the famous adaptive radiation seen in cichlid fishes from Lake Malawi. The authors use structural and sequence information from species belonging to distinct ecotypic categories, representing subclades of the radiation, to document structural variation across the evolutionary tree, infer introgression of inversions among branches of the clade, and even suggest that certain rearrangements constitute new sex-determining loci. The insight is intriguing and is likely to make a substantial contribution to the field and to seed new hypotheses about the ecological processes and adaptive traits involved in this radiation.

I think the paper could be clarified in its prose, and that the discussion could be more informative regarding the putative roles of the inversions in adaptation to each ecotypic niche. Identifying key, large inversions shared in various ways across the different taxa is really a great step forward. However, the population genomics analysis requires further work to describe and decipher in a more systematic way the evolutionary forces at play and their consequences on the various inversions identified.

The model of evolution involving multiple inversions putatively linking together co-adapted "cassettes" could be better spelled out since it is not entirely clear how the existing theory on the recruitment of inversions in local adaptation (e.g. Kirkpatrick and Barton) operates on multiple unlinked inversions. How such loci correspond to distinct suites of integrated traits, or not, is not very easy to envision in the current state of the manuscript.

This is a very interesting point, and we agree creates complications for a simple model of local adaptation. We imagine though that the actual evolutionary history was much more complicated than a single Rhamphochromis-type species separating from a single Diplotaxodon-type species and could have occurred sequentially involving multiple species that are now extinct. A better understanding of the role each of these inversions play in phenotypic diversity could potentially help us determine if different inversions carry variation that could be linked to distinct habit differences. We have added a line to the discussion.

The role of one inversion in sex determination is apparent and truly intriguing. However, the implication of such locus on ecological adaptation is somewhat puzzling. Also, whether sex determination loci can flow across species via introgression seems quite important as a route to chromosomal sex determination, so this could be discussed further.

Another very interesting point. If the inversions are involved in ecological adaptation (an important caveat), then potentially the inverted and non-inverted haplotypes play dual roles in the Aulonocara animals with the inverted haplotype carrying adaptive alleles to deep water and the non-inverted haplotype carrying alleles resolving sexual conflict. We have broadened our discussion about their function at the origin including non-adaptive roles.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

Overall, the paper is well-written and clear. I do have a few suggestions for changes that would help the reader:

(1) Figure 1: the figure legend could be expanded here to help the reader; what are the blue and yellow lines? Why are there two lines for the GT3a assembly? And, I had to somehow read the legend a few times to understand that the top line is the UMD2a reference assembly, and the next line is the new Bionano map.

Fixed in what is now Figure 2

(2) Paragraph starting on line 133: you use the word "test" to refer to the Bionano analyses; it is not clear whether anything is being tested. Perhaps "analyse the maps" or just "map" would be more clear? Or more explanation?

The text has been modified to address this point

(3) L145-146: perhaps change "a single inversion" and "a double inversion" to "single inversions" and "double inversions".

The text has been modified to address this point

(4) L157: suppression of recombination in inversion heterozygotes is "textbook" material and perhaps does not need a reference. Or, you could reference an empirical paper that demonstrates this point. Though I love the Kirkpatrick and Barton paper, it certainly is not the correct reference for this point.

The Kirkpatrick reference was incorrectly included here. The correct reference was an empirical demonstration (Conte) that there were regions of suppressed recombination that have been observed in the location of the inversions. We have also moved this reference further up in the sentence to a more appropriate position

(5) L173: how do you know this is an assembly error and not polymorphism?

The text has been modified to address this point

(6) L277(?): "currently growing in the lab" is probably unnecessary.

The text has been modified to address this point

(7) L298: "the inversion on 10 acts as an XY sex determiner": the inversion itself is not the sex determination gene; rather, it is linked. I think it would be more precise, here and throughout the paper, to say that these inversions likely harbor the sex determination locus (for example, the wording on lines 369-370 is misleading).

We agree with the larger point that the inversion might not be causal for sex determination, however, it could still be causal through positional effects. We have modified the text to make it clear that it could also carry the causal locus (or loci).

(8) Figure 6: overall, this figure is very helpful! However, it contains several problematic statements. In no case do you have evidence that these inversions are "favored by selection"; such statements should be deleted. Also, in point 3, you state that inversions 9, 11, and 20 are transferred to benthic lineages, and then that these inversions are involved in sex determination. But, your data suggests that it is chromosomes 9, 10, and 11 that are linked to sex determination.

This figure is now Figure 1. We have remove these problematic statements.

(9) L356-360: I would move the references that are currently at the end of the sentence to line 357 after the statement about the previous work on hybridization. Otherwise, it reads as if these previous papers demonstrated what you have demonstrated in your work.

The text has been modified to address this point

(10) Overall, the discussion focuses completely on adaptive explanations for your results, and I would like to see at least an acknowledgement that drift could also be involved unless you have additional data to support adaptive explanations.

We have rewritten the text to account for the possibility of drift (line 404 and 405).

Reviewer #3 (Recommendations for the authors):

The paper utilizes heterogeneous datasets coming from different sources, and it is not always clear which specimens were used to generate structural information (bionano) or sequence information. A diagram summarizing the sequence data, methodologies, and research questions would be beneficial for the reader to navigate in this paper.

Much of this information has been added to what is now Figure 1. All of this data is also found in Table S2.

The authors performed genome alignments to analyze and homologize inversion, but this process is not clearly described. For the PCA, SNP information likely involves mapping onto a common reference genome. However, it is not clear how this was achieved given the different species and varying divergence times involved.

We now include a link to the github that contains the commands that were run. Because the overall level of sequence divergence between cichlid species is quite low (2*10^-3 – Milansky et al), mapping different species onto a common reference is commonly performed in Lake Malawi cichlids.

The introgression scenario is very intriguing but its role in local adaptation of the ecogroup types is not easy to understand. I understand this is still an outstanding question, but it is unclear how the directionality of introgressions was estimated. This can be substantiated using tree topology analysis, comparative estimates of sequence divergence, and accumulation of DNA insertions. The diagram does not clearly indicate which ones are polymorphic. In some cases, polymorphic inversions could result from the coexistence of native and introgressed haplotypes.

We agree that this analysis would be interesting but is beyond the scope of this paper.

The alternative model of introgression proposed in the cited preprint is interesting and should deserve a formal analysis here. The authors consider unclear what would drive "back" introgressions of non-inverted haplotypes, but this would depend on the selection regimes acting on the inversions themselves, which can include forms of balancing selection and a role for recessive lethals (heterozygote advantage). For instance, a standard haplotype could be favored if it shelters deleterious mutations carried by an inversion. Testing the introgression history over a wider range of branches and directions would provide further insights.

We agree that this analysis would be interesting but is beyond the scope of this paper.

The prose in the paper is occasionally muddled and somewhat unclear. Referring to chromosomes solely by their numbers (e.g.. "inversion on 11") complicates readability.

This is the standard way to refer to chromosomes in cichlids and we believe while it complicates readability, any other method would be inconsistent with other papers. Changes to nomenclature might improve the readability of this paper, but would make it more difficult to compare results for these chromosomes from other papers with what we have found.

Version 2

Abstract

Abstract

Chromosomal inversions are an important class of genetic variation that link multiple alleles together into a single inherited block that can have important e7ects on fitness. To study the role of large inversions in the massive evolutionary radiation of Lake Malawi cichlids, we used long-read technologies to identify four single and two tandem inversions that span half of each respective chromosome, and which together encompass over 10% of the genome. Each inversion is fixed in one of the two states within the seven major ecogroups, suggesting they played a role in the separation of the major lake lineages into specific lake habitats. One exception is within the benthic sub-radiation, where both inverted and non-inverted alleles continue to segregate within the group. The evolutionary histories of three of the six inversions suggest they transferred from the pelagic Diplotaxodon group into benthic ancestors at the time the benthic sub-radiation was seeded. The remaining three inversions are found in a subset of benthic species living in deep waters. We show that some of these inversions are used as XY sex-determination systems but are also likely limited to a subset of total lake species. Our work suggests that inversions have been under both sexual and natural selection in Lake Malawi cichlids and that they will be important to understanding how this adaptive radiation evolved.

Introduction

Large genomic inversions are a particularly interesting type of genetic variation in the evolutionary process1–4. These rearrangements of DNA sequence strongly suppress recombination between inverted and non-inverted alleles, enabling the capture and accumulation of genetic variants of smaller molecular size that are linked together on each inversion haplotype5. These alternative haplotypes largely follow independent evolutionary paths within a species, accumulating new mutations that can aMect phenotypes under natural and sexual selection. Structural rearrangements can be large, spanning multiple megabases of DNA containing hundreds of genes2,6. They have been shown to play a role in adaptive divergence within a variety of species7–12, in creating alternative reproductive strategies and mating types13–16, in evolution of sex chromosomes1,17, and in the formation of pre- and post-zygotic barriers between incipient species18,19.

When comparing the genomes of extant species, chromosomal inversions are often found as fixed diMerences, consistent with a role in speciation20–22. Inversions have been shown to play a role in adaptation to diMerent environments and the establishment of prezygotic and postzygotic barriers21,23. However, the evolutionary forces responsible for establishing and spreading inversions remain obscure. Natural selection and sexual selection are thought to play a role but neutral forces like genetic drift can also be responsible for their spread. Perhaps most likely, all of these diMerent evolutionary forces contribute to their fixation24,25.

Few vertebrates oMer as much potential for investigating evolution and speciation as the Cichlidae family of fish, one of the most species-rich and diverse families of vertebrates, with an estimated 2,000-3,000 species across the globe 26–28. Evolutionary radiations have occurred at least three times in the African Great Lakes (Lake Malawi, Lake Tanganyika, and Lake Victoria). In Lake Malawi, more than 800 species of cichlids evolved in the past 1.2 million years27,29–31. Speciation in cichlid fishes involved extremely high levels of phenotypic divergence, including changes in body shape, jaw structures, and feeding behaviors related to prey in their ecological niches26,32. Additionally, traits that create prezygotic barriers between species are also extremely diverse. Color patterns are quite dramatic and diverse among species and include both diMerences in color and species-specific patterns including bars, stripes, and blotches that are thought to reinforce barriers between species in sympatry33. In some species, mating is seasonal, with gatherings on ‘leks’ where males build bowers by scooping and spitting out sand over the course of a week or more34. Females choose mates based upon features of these bowers including shape and size35. Most of the Lake Malawi species can be induced to interbreed in the lab, providing the opportunity to use genetic approaches to study causality of associated changes. Despite the presence of these prezygotic barriers, substantial gene flow has occurred multiple times within the lake30,36.

Our current understanding of how the Lake Malawi radiation created >800 species is a set of three serial diversifications, starting from a single riverine-like ancestor lineage, that created seven major ecogroups that currently live in the lake30 (Figure 1). Each ecogroup forms its own clade in a whole-lake phylogeny and occupies distinct habitats within Lake Malawi. A pelagic lineage separated first, further diversifying into a deep-water ecogroup (Diplotaxodon) and a mid-water ecogroup (Rhamphochromis). A muddy/sandy benthic lineage evolved from the ancestral riverine lineage next, which split into three ecogroups (collectively referred to as benthics) living either in the water column (utaka), over shallow sandy/muddy shores (shallow benthics), or in deep-water habitats (deep benthics). Finally, an ecogroup living over rocky habitats evolved (mbuna). The riverine generalist Astatotilapia calliptera (AC) living in the border regions of the lake and surrounding rivers, is the seventh ecogroup, and is thought to represent the ancestral lineage that seeded the three primary Lake Malawi lineages. Subsequent radiations within these ecogroups based on trophic specializations and sexual selection further diversified the species flock30,32.

Overview of Lake Malawi cichlids.

A summary of the Lake Malawi cichlids and the results from this paper. On the left, a phylogeny showing the seven major lineages that currently live in the lake which are referred to as ecogroups. AC is Astatotilapia calliptera, a riverine species that lives in the margins of the river/lake interface. For each ecogroup, we summarize the 1) estimated number of species in the lake, 2) the number of species we have Bionano optical read data for, 3) the number of species we have PacBio based genome assembies, 4) the number of species with Illumina short-read sequencing data, and 5) the primary habitat these ecogroups live in. Deep and shallow benthics are distinguished by the lake bottom depths where they live while utakas live in the water column but mate over the shore. We also summarize the distribution of 6 large inversions we have identified, colored by the chromosome they are found on (legend). We propose that the inversions on 9, 11, and 20 originated in the pelagic lineage and spread to the benthics by hybridization very early on in the evolution of the benthic lineage. Alternatively, these inversions distribution could be explained by incomplete lineage sorting.

Here, we investigate the role large inversions played in the Lake Malawi radiation using three technologies. Using the Bionano Saphyr optical mapping system, we identify four single large inversions and two double inversions, ranging from 9.9Mb to 20.6Mb in size. A fifth single inversion, composed of just one of the two tandem inversions on chromosome 20, was also identified, indicating a serial set of rearrangements created this structural variant. We validate these inversions using new genome assemblies created using PacBio Hifi reads. Finally, we use existing and newly generated short-read Illumina sequencing to genotype these inversions in DNA samples from cichlids collected from Lake Malawi. These inversions primarily segregate by ecogroup, with no inversions found in the mbuna and AC samples, one inversion fixed in Rhamphochromis, three inversions fixed in Diplotaxodon, and all six segregating within the benthic sub-radiation. The evolutionary histories of these inversions are inconsistent with the species phylogeny and suggest that the three Diplotaxodon inversions spread into the benthics via hybridization around the time the benthic lineage diverged from the ancestral riverine species. The three additional inversions are found primarily in benthic species living in deep water habitats. We provide evidence that three of the inversions are involved in sex determination in a subset of benthic species. Our work provides a framework for understanding the role of inversions in the adaptive radiation of Lake Malawi cichlids, and supports future investigations into their roles in the establishment of ecogroups, adaptation to deep water habitats, and in controlling traits under sexual selection.

Results

Improvement of the Metriaclima zebra reference

The mbuna cichlid, Metriaclima zebra, was the first species sequenced as a reference genome for Lake Malawi cichlids37. The current version of its genome, M_zebra_UMD2a, has been created by assembling PacBio reads into contigs which were anchored to 22 chromosomes using recombinant maps generated using crosses between diMerent Lake Malawi species38. While the overall quality of this reference is high compared to other non-model organisms, a total of 20.5% of the DNA was not anchored to a chromosome. Additionally, the orientation of contigs can be ambiguous given how contigs were anchored to linkage groups using recombinant maps. These imperfections could complicate inversion discovery, so we first sought to improve the reference sequence.

The Bionano Saphyr system generates large (150 kb - multiple Mbs) DNA molecules and labels a specific motif (CTTAAG) with a fluorescent probe39. While these molecules do not provide sequencing information, the distance between observed fluorescent loci can be used to assemble multiple molecules into large maps, with lengths spanning up to an entire chromosome. Comparison of these maps with an in-silico digest of the reference sequence identifies large structural changes in the observed sample or errors in the reference genome. We processed DNA collected from the blood of three Metriaclima zebra individuals using the Bionano pipeline and assembled DNA molecules into optical maps. We identified many discrepancies between the observed optical maps and the M_zebra_UMD2a assembly (Figure 2). These diMerences fall into two primary categories: 1) contigs that were incorrectly ordered or oriented and 2) unplaced contigs that Bionano anchored to specific chromosomes. The first category is consistent with errors in the assembly caused by ambiguity in placing contigs into a linkage group using recombination maps. Our results indicate that 178 of the contigs are incorrectly oriented and 133 of the unanchored contigs could be placed onto specific chromosomes.

Generation of a new high quality reference genome for Metriaclima zebra.

Alignment of Bionano maps derived from a single Metriaclima zebra male to two different reference sequences for Chromosome 17 (white and grey, respectively). M_zebra_UMD2a reference (white) is the existing reference genome and M_zebra_GT3a reference (grey) was created for this manuscript using a combination of PacBio HiFi reads and Bionano molecules. For each reference, the top bar shows the expected location of each CTTAAG motif (blue) created from an in silico annotation of the reference sequence. These are aligned (grey lines) to observed locations in maps generated from Bionano Saphyr optical molecules (bottom bar). Deviations from co-linearity indicate reference errors (green, blue, and red highlights). Yellow indicates regions of the genome that lack aligned motifs. A summary of the improvements in the new genome is found in Table S1.

To further improve the reference, we used PacBio HiFi reads generated from an additional male Metriaclima zebra individual to create a new genome assembly, taking advantage of the low overall error rate (< 0.1%) of HiFi reads compared to previous generation PacBio technologies. Contigs generated from these data were combined with Bionano maps to generate a new hybrid genome assembly, which we refer to as the M_zebra_GT3a assembly. The overall quality of the assembly was excellent, with 933Mb of 962 Mb of DNA assigned to the 22 linkage groups, (Table S1) reducing the amount of unplaced contigs from 196 Mb to 28 Mb. The N50 length was 32.542Mb, with many of the scaMolds containing entire chromosomes.

Identification of six large inversions from 11 species of Lake Malawi cichlids

Armed with the improved reference, we used the Bionano system to test 31 samples from 11 diMerent species found in Lake Malawi. Eight of the species were bred in the Streelman or McGrath laboratories (Table 1 and Table S2). These included four mbuna species (Pseudotropheus demasoni, Cynotilapia zebroides, Labeotropheus fuelleborni, and Labeotropheus trewavasae), two shallow benthic species (Mchenga conophoros and Nyassachromis prostoma “Orange Cap”), a single deep benthic species (Aulonocara sp. ‘chitande type north’ Nkhata Bay), and a single utaka species (Copadichromis virginalis). To expand our coverage of the Lake Malawi radiation, we also obtained samples from the pet trade for a single species of the Diplotaxodon ecogroup, Diplotaxodon limnothrissa, and a single species of the Rhamphochromis ecogroup, Rhamphochromis longiceps. Finally, we also tested a single male Protomelas taeniolatus shallow benthic individual collected from Lake Malawi based on its position in principal component analysis of SNVs described below. In total, we were able to obtain samples from 6 of the 7 ecogroups, excluding Astatotilapia calliptera.

From these 31 individuals, we identified six new inversions ranging from 9.9 - 22.9 Mb in size (Figure 3): single inversions on chromosomes 2, 9, 10, and 13, and double inversions on chromosomes 11 and 20. Interestingly, regarding the double inversion on chromosome 20, we also identified four Rhamphochromis individuals that carried only the first inversion, lacking the 4.1 Mb long second part of this rearrangement (Figure 3). This indicates that the tandem inversions on 20 were formed by at least a two-part process, most parsimoniously with the left arm of the inversion (20a) forming first followed by the right inversion (20b) occurring in that genetic background.

Identification of six large single or double inversions segregating within Lake Malawi cichlids.

Representative alignments of inversions or double inversions identified from blood samples obtained from 30 individuals from eleven species. For each inversion, the top shows predicted CTTAAG motifs from an in-silico annotation of the reference genome, the bottom shows motifs identified using Bionano molecules obtained from an individual of the species indicated on the left, and the grey lines indicate matching motifs based upon predicted and observed distances. Yellow indicates regions containing no matches between motifs in the reference genome and the mapped sample. Single inversions were identified on chromosomes 2, 9, 10, 13, and 20. Tandem double inversions were identified on chromosomes 11 and 20. The estimated length and percentage of the chromosome spanned by the inversion is shown on the right. The single inversion on 20 identified in Rhamphochromis longiceps has the same position as the first inversion of the double inversion on 20 identified from Mchenga conophoros. A list of all samples and their inversion genotype is in Table 1. Note that an error or polymorphism in the reference genome in chromosome 2 is indicated in the first panel.

The distribution within these species is summarized in Table 1. For simplicity, we will refer to all six rearrangements as inversions, regardless of whether they are single or double inversions. The most common inversions were on chromosome 9, 11, and 20 which were found in a combination of benthic and Diplotaxodon individuals. The position of these three inversions were also consistent with regions of low recombination rate previously identified in an intercross between a mbuna and benthic individual38, as expected, since inversions suppress recombination between inverted and non-inverted haplotypes. The remaining three inversions (2, 10, and 13) were only found in the deep benthic individuals (Aulonocara sp. ‘chitande type north’ Nkhata Bay). While for most samples the inversion genotype was homozygous, for five individuals the optical maps supported both inverted and non-inverted haplotypes, indicating the inversions were heterozygous in those samples (Figure S1).

de novo reference genomes support the presence of these six inversions

To provide further support for these structural rearrangements, we generated de novo hybrid genome assemblies of three Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals, which carry all six of the identified inversions, and two additional Metriaclima zebra individuals using a combination of PacBio HiFi reads and Bionano maps. The overall quality of these assemblies is high, with large N50 values and a small number of individual contigs (Table S3). We performed whole genome alignments between the five new

assemblies and the GT3a reference genome (Figure S2-S7). These alignments show that the new assemblies support the presence of all six inversions in the Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals. For most inversions, a single scaMold covered the entire inversion with breakpoints in locations consistent with the Bionano maps. The two new control assemblies were consistent with the M_zebra_GT3a reference and did not show any evidence of inversions in those regions. In other regions of the genome, however, there were diMerences in the alignment of the M_zebra_GT3a assembly and the Metriaclima zebra genomes, most prominently in the left arm of chromosome 2 (Figure 3) and in the highly repetitive chromosome 3, which likely indicates some errors in the M_zebra_GT3a assembly or that a large polymorphism segregates within the species.

We refined the location of the end points for each inversion using the genomic alignments. However, the presence of a large amount of repetitive DNA in regions near these regions prevented us from defining inversion boundaries at single base pair resolution. The mechanisms that create large inversions often involve repetitive sequence near the breakpoints, making exact localization diMicult if not impossible40.

We also took advantage of a genome assembly for a Rhamphochromis sp ‘chilingali’ individual (fRhaChi2.1) created by the Darwin Tree of Life Project produced using PacBio data and Arima2 Hi-C data (Accession PRJEB72870). The genome alignment between the fRhaChi2.1 assembly and the GT3a assembly confirmed a single inversion on chromosome 20, consistent with the Bionano data (Figure S2- S7).

For each inversion, we determined the ancestral haplotype using two species as outgroups, Oreochromis niloticus and Pundamilia nyererei. Oreochromis niloticus, or Nile tilapia, is an important food fish that’s estimated to have diverged from Lake Malawi cichlids 14.1 to 30 million years ago while Pundamilia nyererei is a more closely related haplochromine found in Lake Victoria37,41. For this analysis we used a published whole genome assembly for Oreochromis niloticus (UMD_NMBU)38 and generated a hybrid genome assembly for Pundamilia nyererei using a previously published whole genome assembly37 combined with Bionano maps we produced from a single individual (Figure S8). For all six inversions, the reference haplotype found in the GT3a assembly represents the ancestral state.

Inference of the segregation of these inversions in Lake Malawi using short-read sequencing

The distribution of these inversions within the lake can provide important information to their function. Genotyping inversions using short read sequencing is possible because each haplotype is on an independent evolutionary trajectory due to the suppression of recombination. Population genetics approaches take advantage of the fact that inversions create patterns of inheritance between SNVs that are readily detected by approaches such as Principal Component Analysis (PCA)42.

We analyzed short read sequencing data from the 31 samples with Bionano data along with 297 wild individuals published by the Durbin lab30, 15 wild individuals published by the Streelman lab 31,43, and 22 wild individuals sequenced for this paper, aligning reads and calling small variants against our new M_zebra_GT3a reference genome. PCA analysis was used to analyze the major axes of variation across the whole genome and within the six inversions. We separately analyzed the left and right inversions on 11 and 20. For the whole genome, the species within an ecogroup clustered together within the first two principal components, reflecting their evolutionary history. This matched the previously reported results of Malinsky et al30. Individuals from four of the ecogroups - Rhamphochromis, Diplotaxodon, Astatotilapia calliptera, and mbuna - formed four distinct clusters (Figure 4A and Figure S9). A fifth, more diMuse cluster, composed of shallow benthics, deep benthics, and utakas, was also present. The broader distribution of the benthic ecogroups in PC space is consistent with both their recent separation as well as large amounts of gene flow thought to have occurred between these diMerent species30,36.

Distribution of inversions within the Lake Malawi ecogroups.

Principal component analysis was used to analyze SNVs distributions in 365 samples within each inversion to genotype the inversion haplotype for all six inversions. (A) We illustrate the approach for genotyping the chromosome 10 inversions. The PCA plot for the entire genome is shown on the left, with each sample colored by its ecogroup. Samples from each ecogroup cluster together, reflecting their evolutionary history. Individuals with Bionano data are shown as grey Xs. On the right, for the SNVs that fall within the chromosome 10 inversion, the deep benthic/shallow benthic and utaka individuals split into multiple clusters. Using the samples genotyped with Bionano data, we assign each cluster to an inversion genotype, which is shown by the black, green, and red boxes. To make comparisons of the whole genome PCA and chromosome 10 PCA plots easier, we reversed the y-axis on the right panel. The assigned genotypes of each sample can be found in Supplemental Table 2. Interactive PCA plots for each inversion are included in Figures S9-S17. (B) The derived allele frequency was calculated for each inversion within the seven ecogroups. The number of species that were genotyped and the estimated number of species that live in Lake Malawi are listed in the whole-genome phylogeny (left).

The PCA plots using the SNVs within the six inversions often violated the pattern of the whole genome PCA (Figure 4A and Figure S10-S17). Using samples that were sequenced with both Illumina and Bionano technologies, we assigned each cluster to specific inversion genotypes (Table S2). We did not have Bionano data for two clusters (one on chromosome 9 and one on 11 – see Figures S11, S13, and S14), so we supplemented our dataset by adding Bionano data from an additional shallow benthic individual (Protomelas taeniolatus) that fell into both clusters. The PCA clusters varied in their complexities for each inversion depending largely on their distribution within the benthic ecogroup (more on this below) and included both homozygous genotypes and heterozygous genotypes. We summarize the overall distribution of the inversions within each of the ecogroups in Figure 4B.

All specimens of the mbuna and AC ecogroup were fixed for the homozygous non-inverted genotype for all six inversions. In Rhamphochromis, all the inversions were homozygous non-inverted except for the left arm of 20 which was fixed for the single inversion state. The Diplotaxodon individuals all carried the non-inverted alleles for the inversions on 2, 10 and 13 and were fixed for the inversions on 9, 11, and 20.

The distribution of these inversions was the most complicated within the benthics. The inversions on 2, 10, and 13, which were identified in Aulonocara sp. ‘chitande type north’ Nkhata Bay, were mostly found together in other deep benthics, including the Alticorpus (geo7reyi, macrocleithrum, peterdaviesi), Aulonocara (blue chilumba, gold and minutus), and Lethrinops (gossei, longimanus ‘redhead’, and sp. olivera) genera. The inversions on 2 and 13 were also found in the shallow benthic species Placidochromis longimanus. However, not all the deep benthics carried the three inversions, including five Aulonocara species (baenschi, gertrudae, steveni, stuartgranti, and stuartgranti Maisoni) and a single Lethrinops species (longipinnis). The distribution of these three inversions in benthic species known to inhabit the deepest habitats are suggestive for a role in depth adaptation.

Both inversions on 20 were homozygous inverted in all benthic individuals. For the single inversion on 9, many benthics were homozygous for the inverted haplotype. However, an additional PC cluster was present, composed of 64 shallow benthic and utaka individuals heterozygous for the non-inverted haplotype (Figure S1). Interestingly, no benthic individuals were identified as homozygous for the non-inverted state.

Finally, the distribution of the chromosome 11 inversion was also determined. Individuals of all three genotypes (homozygous non-inverted, homozygous inverted, and heterozygous) were observed (Figure 4a).

Evolutionary history of 9,11, and 20 suggest introgression from Diplotaxodon into benthic ancestors

The distribution of these inversions conflicts with the species tree, suggesting that these inversions could have spread to the benthics via hybridization or carried forward by incomplete lineage sorting. To determine their evolutionary history, we built phylogenies for the whole genome as well as for each inversion using SNVs that fell within each of the six inversions using representative species from each ecogroup (73 total – Table S2) as well as two previously sequenced Pundamilia nyererei individuals as an outgroup (Figure 5A and Figures S18 – S25)44,45. To avoid issues with haplotype phasing, we focused on individuals that were homozygous for one of the two inverted genotypes.

Evolutionary history of large inversions violates species tree.

(A) Maximum likelihood trees built from either whole genome SNVs (top) or within the 9, 11, and 20 inversions. Between 2-25 representative samples were selected for each ecogroup (see Table S2) and individuals that were heterozygous were excluded from this analysis. Clades were collapsed by ecogroup for visualization purposes. Full phylogenies are presented in Figures S18 to S25. The whole genome tree indicates Diplotaxodon and Rhamphochromis species split first and formed their own clade. For each of the displayed inversions, the first split was between samples carrying the inverted vs. non-inverted genotypes. Additionally, the Diplotaxodon individuals now formed a clade with the benthics that carry the inversion. (B) Density plots of dXY values comparing benthic species and species of the indicated ecogroup. For the displayed inversions, the Diplotaxodon species were much closer in evolutionary distance when compared to the whole genome plot.

While the whole genome phylogeny matched the expected topology based upon our understanding of the Lake Malawi species tree, the topologies of each of the inversions showed key diMerences between the separation of ecogroups. The benthic group often split into separate clades, correlated with their inversion genotype. Prominently, we found the three inversions on 9, 11, and 20 showed key diMerences in the relationship between the Diplotaxodon and benthic ecogroups. While Diplotaxodon is a sister group to Rhamphochromis at the species level, for these three inversions, the Diplotaxodon individuals formed a clade with the benthics that also carried the inversion. Additionally, genetic distance between Diplotaxodon and benthics carrying the inversion was also much smaller than the rest of the genome (Figure 5B). Both the phylogenies and genetic distances between ecogroups are consistent with the introgression of the 9, 11, and 20 inversions from the Diplotaxodon to benthic ancestors at the time of the benthic radiation.

A role for large inversions in sex determination within the benthic ecogroup

The evolution of sex determination is often associated with the presence of inversions to repress recombination between the sex chromosomes46,47. We tested whether these inversions could play a role in sex determination in benthic species, as we identified individuals that were heterozygous for some of these inversions (Table 1). We started with the inversion on 10, as our Bionano data identified three male Aulonocara sp. ‘chitande type north’ Nkhata Bay individuals that were heterozygous and a single female that was homozygous for the inverted haplotype. To test whether this inversion is linked to a genetic locus (or loci) that act as an XY system in this species, we tested two separate broods (Brood 1; 2 parents, 24 oMspring and Brood 2: 2 parents, 24 oMspring) currently growing in the laboratory, sequencing 12 males and 12 females from each brood with Illumina paired-end sequencing. We used PCA to genotype each of the 52 animals (4 parents and 48 oMspring) for the chromosome 10 inversion genotype. The association in the oMspring between the inversion 10 genotype and sex was almost perfect: 24 females and 1 male were homozygous for the inversion, while 23 males were heterozygous for the inversion (Figure 6B). These individuals were sexed by venting so the imperfect segregation could either indicate an error in sexing or that other modifiers influence sex.

A role for the chromosome 10 inversion in sex determination

(A) Schematic of the two inversion states (normal and inverted) and their assignment as X or Y chromosomes. (B) 24 offspring (12 male and 12 female) from two separate male/female pairs were genotyped for the inversion. For 47 of 48 individuals, the XY animals were male and the XX animals were female. One male was XX. (C) FST analysis comparing 24 males and 24 females identified chromosome 10 as the primary genetic region that segregated with sex. (D) Comparison of heterozygosity levels of 24 male and 24 female animals also identified chromosome 10 as an outlier.

We also used binned Weir Cockerham Fst analysis, grouping male and female individuals to scan the genome for regions that were associated with sex (Figure 6C). The inversion on 10 was the region of the genome that was in strongest linkage disequilibrium with sex. Males also had a much higher heterozygosity in this region than females (Figure 6D). Altogether, these results indicate that the inversion on 10 acts as an XY sex determiner in the Aulonocara sp. ‘chitande type north’ Nkhata Bay species.

We also tested eight laboratory Nyassachromis prostoma “Orange Cap” individuals, as our Bionano data identified a male that was heterozygous for the inversion on 11 and a female that was homozygous for the inversion. We performed short-read sequencing on three males and five females growing in the lab, genotyping the inversion using PCA analysis. All three males were heterozygous for the inversion while the five females were homozygous, a significant association between genotype and sex (Fisher exact test p- value = 0.0179). This suggests a separate XY system segregates in Nyassachromis prostoma “Orange Cap”, linked to the inversion on chromosome 11.

Discussion

Here, we characterize the presence of large inversions found in the Lake Malawi cichlid flock, identifying six inversions segregating within the lake, providing a framework to address the role structural variants played in this lake’s adaptive radiation (Figure 6). What are the evolutionary forces responsible for their spread and what role do they currently play in the lake? Inversions have multiple, well-characterized roles in adaptive evolution: 1) they can capture beneficial alleles responsible for adaptation to local habitats, 2) they can capture alleles that create prezygotic and postzygotic barriers that prevent gene flow during speciation, and 3) they can capture alleles that influence sex-specific phenotypes that are linked to sex-determination systems1,3,4. These inversions could have also spread through neutral processes like genetic drift or through other adaptive processes not listed here.

This question is much more complicated in the context of the Lake Malawi cichlid radiation with ∼800 diMerent species. The role of these inversions must be considered in a species- and time-specific way. The recent history of the lake was quite dynamic, with large cyclical changes in water depth and shoreline in the past 800,000 years that are predicted to create alternating periods of extinctions and radiations48–50. Similarly, the role the inversions play in each species (such as sex determination) will not be the same. Determining whether these inversions carry alleles that contribute to the phenotypic diversity found in Lake Malawi will be an important first step towards resolving the evolutionary forces responsible for their spread through the lake. This is possible as most species in the lake can interbreed, allowing for quantitative genetics approaches to link genotype variation to phenotypic variation and determine what traits are influenced by the inversions.

The local adaptation mechanism is one well-characterized method to explain how inversions can spread through adaptive evolution6,8. Under this hypothesis, two populations undergoing adaptive divergence in the presence of migration and gene flow create conditions that can select for inversions, provided they capture and link adaptive alleles together that increase fitness to their local environments51. Apart from the benthics, the inversions are fixed in each ecogroup in the inverted or non-inverted state, each of which live in unique habitats in the lake that place very diMerent evolutionary pressures on the animals that live within them. In this scenario, the inversions would capture adaptive alleles that increase fitness of ancestral species to pelagic and deep-water environments. To illustrate this scenario, we consider the time the pelagic groups split from the ancestral AC fish that is thought to live in rivers and the borders of the lake (Figure 1). If the first inversion on 20 captured alleles that allowed these pelagic ancestors to inhabit the lake, it could be selected on provided that these pelagic ancestors continue to interbreed with the AC ancestors. Similarly, the diversification of the Rhamphochromis and Diplotaxodon groups could have involved the inversions on 9, 11, and right arm of 20, and the adaptation of benthics to deeper water habitats could have involved the inversions on 2, 10, and 13. We do note that the existing theory for recruiting inversions to local adaptation does not account for multiple unlinked inversions (i.e. 9, 11, and right arm of 20). Future experimental and comparative genomics approaches can determine if local adaptation was at play at the time these lineages separated.

We also provide evidence that two inversions play a role in sex determination in the benthics, with the strongest evidence showing the inversion on 10 acts as an XY system in a deep benthic species. Additionally we show that the inversion on 11 is significantly associated with sex determination in a shallow benthic species as an XY system. Despite the importance of the maintenance of distinct, reproductively compatible sexes, sex determination systems can be evolutionarily labile, and Lake Malawi cichlids show exceptional variability, with at least 5 diMerent genetic sex determination systems described to date52,53. While we implicated two inversions in sex determination, our work also underscores how dynamic sex determination is within the benthics. While many benthic species carry the diMerent inversions, their genotypic distribution in most species is inconsistent with a role for sex determination. EMectively, the genetic determination of sex determination largely must be determined in a species by species way, and individual species may be segregating multiple sex determination systems. The selective forces on sex chromosomes are strong, including balancing selection for sex ratio and sexually antagonistic selection on sexually dimorphic phenotypes. Lake Malawi cichlids show amazing sex diMerences, characterized by diversification in numerous phenotypes, including pigmentation, behavior, and morphology thought to contribute to species barriers. The use of inversions in sex-determination allows for the accumulation of alleles in linkage with the sex-determiner gene which will have sex-specific eMects in the male individuals that carry the Y chromosomes. The presence of these inversions in various ecogroups could simply be remnants of their role in sex determination in the ancestors of pelagic ecogroups, and have nothing to do with local adaptation, which could have contributed to their spread before sex chromosome turnover ended balancing selection on the inverted/non-inverted haplotypes and subsequent fixation of these inversions. Identification of the causal alleles for sex determination on these inversions will help us to resolve this question.

The presence of non-inverted haplotypes in the benthics could have been retained from the time the lineage initially formed, with both haplotypes under balancing selection if they were used for sex determination. A recent preprint, however, which also identified these 6 inversions presented here using multiple technologies, also utilized short-read sequencing of 1375 individuals to create an alternative model54. They inferred the non-inverted haplotypes of the 9 and 11 inversions were introgressed back into the benthics from riverine species within and outside of the lake. Similarly, the non-inverted haplotypes on chromosomes 2, 10, and 13 in the deep benthics were introgressed from shallow benthics. In this model, it is unclear what evolutionary forces drove these back introgressions of the non-inverted alleles, but would imply these inversions have not continuously provided the genetic sex determination mechanism since their inception.

The distribution of the inversions in the benthic sub-radiation is uniquely complicated in Lake Malawi. Inverted and non-inverted haplotypes for five of the six inversions are segregating within this group. Additionally, the presence of the 9, 11, and 20 inversions within benthics violates the whole genome phylogeny. We propose these three inversions originated within the Diplotaxodon group and spread to a very early benthic ancestor via introgression. We favor this over incomplete lineage sorting, which would also lead to violations of the phylogeny, for a few reasons. First, the dynamic history of the lake levels is consistent with times of extinction and hybridization48–50. Second, genomic analysis of extant cichlids provided evidence for a large amount of hybridization events36,55. Finally, the close genetic distance between Diplotaxodons and benthics that carry the inverted haplotype (Figure 5) is consistent with hybridization. Hybridization has been proposed to fuel adaptive radiations of cichlids in all three great African lakes (Malawi, Tanganyika, and Victoria) 27,28,56 and our work suggests that an ancient hybridization event between early benthic and Diplotaxodon ancestors occurred prior to the benthic radiation. It will be interesting to study the role of these three inversions in benthic species. Diplotaxodon are pelagic, deep-water dwellers that feed on plankton or small fish while benthics live near the shore over sandy and muddy habitats30. It is not obvious how the traits of the Diplotaxodon would benefit benthic animals that live in such dissimilar environments. However, the initial habitats that the benthic ancestors utilized is not known and potentially their initial habitat was in deeper waters than the AC ancestors.

This report provides an initial framework for understanding the role of inversions in the adaptive radiation of Lake Malawi cichlids. Future work identifying and characterizing causal alleles captured within these inversions will enable us to discern the specific mechanisms by which they contribute to speciation using this remarkable evolutionary system as a model for the evolution of all animals.

Methods

Samples

Wild-caught cichlids were acquired from Old World Exotic Fish Inc in 2022 and Cichlidenstadel in 2024. Lab-reared species were purchased from Southeast Cichlids, Old World Exotic Fish Inc., and Cichlidenstadel at various points in the past two decades (Table S2). Fish were delivered live to the Georgia Institute of Technology cichlid aquaculture facilities. All samples were collected following anesthetization with tricaine according to procedures approved by the Institutional Animal Care and Use Committee (IACUC protocol numbers A100029 and A100569).

Caudal fins were collected from subjects and flash frozen on powdered dry ice before storage at −80°C. Care was taken to minimize freeze-thaw cycles prior to DNA extraction via Bionano or Qiagen protocols. If necessary, fins were sectioned on a sterile aluminum block set in dry ice prior to digestion.

Blood was collected following rapid decapitation of anesthetized subjects. Wide-bore, low-retention pipette tips, 0.5M EDTA (pH 8.0, Invitrogen Cat# 15575-038), and Bionano Cell BuMer (Part Number 20374) prevented blood from clotting at room temperature. Samples were used immediately for Bionano DNA extraction (see below).

Bionano

Bionano DNA extractions were predominantly performed on fresh whole blood samples (see above). Blood concentration was determined via a manual logical count performed on a hemocytometer and 2 million cells were carried forward for DNA extraction with the Bionano SP-G2 Blood & Cell Culture DNA Isolation Kit (Part Number 80060) as per the Bionano Prep SP-G2 Fresh Human Blood DNA Isolation Protocol (Document Number CG-00005, Rev C). Because nucleated cichlid blood had high concentrations, steps 3 and 4 were skipped and samples were resuspended in Bionano Cell BuMer to a total volume of 200µL. Following extraction, samples were allowed to homogenize undisturbed at room temperature for at minimum 72 hours prior to quantification using the Qubit dsDNA Quantitation, Broad Range kit (ThermoFisher, Cat # Q32853).

Bionano data for a single Protomelas taeniolatus sample (PT_2003_m, see Table S2) was generated from fin tissue using a combination of the Bionano Prep SP Tissue and Tumor DNA Isolation Protocol (Document Number 30339, Rev A) and the Bionano Prep SP-G2 Fresh Human Blood DNA Isolation Protocol.

750ng of homogenized DNA was fluorescently labeled using the Bionano Direct Label and Stain-G2 (DLS- G2) Kit (Part Number 80046, Protocol - CG-30553-1, Rev E). Labeled DNA was quantified using the Qubit dsDNA Quantitation, High Sensitivity kit (ThermoFisher, Cat # Q32854) and loaded onto a flow cell within a Bionano Saphyr Chip G3.3 (Part Number 20440). Samples were imaged on the Bionano Saphyr with at least 500Gbp of data collected per sample.

Molecules for each sample were assembled using the Bionano de novo assembly pipeline on Bionano Access Version 1.7. The preassembly setting was turned on, and the variant annotation parameters were deselected for every assembly. De novo maps were aligned to the Mzebra_GT3a genome.

PacBio HiFi sequencing and genome assembly

PacBio de novo assemblies were generated from HiFi reads across three sequencing runs. The MZ_GT3.3 and YH_GT1.3 were sequenced following DNA extraction from frozen caudal fin tissue using the Qiagen MagAttract HMW DNA Kit (Cat. No. 67563). In the first sequencing run a Metriaclima zebra female was sequenced sequenced on a PacBio Sequel II system by the Georgia Genomics and Bioinformatics core. In the second run, the same the Metriaclima zebra female and an Aulonocara sp. ‘chitande type north’ Nkhata Bay female was sequenced on a PacBio Revio instrument by the HudsonAlpha Institute for Biotechnology. Reads from both runs were used to assemble the MZ_GT3.3 genome and reads from the second run alone were used to assemble the YH_GT1.3 genome (see details below).

Mzebra_GT3a, MZ_GT3.2, YH_GT1.1, and YH_GT1.2 were assembled using HiFi reads generated with DNA extracted from fresh whole blood and heart tissue. DNA was extracted via the PacBio Nanobind Tissue Kit RT (Part Number 102-208-000) following a modified DNA from animal tissue using the Nanobind® PanDNA kit protocol (Part Number 102-574-600). Blood and heart tissue were combined and homogenized together using a Qiagen TissueRuptor (Qiagen 9002755) and centrifugation g-force was halved during the DNA extraction. DNA fragments <25kb were removed using the PacBio Short Read Elimination kit (Part Number 102-208-300). Library preparation and sequencing was performed by the University of Maryland Institute for Genome Sciences on a PacBio Revio instrument.

HiFi reads from all samples were assembled using the Mabs de novo assembler (v2.28)57 using the mabs-hifiasm algorithm and default parameters. Each assembly from mabs-hifiasm was evaluated using Inspector (inspector.py v1.0.1)58 run on default parameters. The assemblies were error corrected with inspector-correct.py (v1.0) to resolve large structural errors. These error-corrected contigs from were uploaded to Bionano Access.

The MZ_GT3.2, MZ_GT3.3, YH_GT1.1, YH_GT1.2 and YH_GT1.3 scaMold-level assemblies were generated by bridging gaps in their contigs using Bionano maps via the single-enzyme Bionano Hybrid ScaMold pipeline using default parameters. A unique set of Bionano maps was used to scaMold the contigs from each de novo assembly. The resulting hybrid scaMold NCBI.fasta file (which denotes gaps filled by Bionano maps with stretches of N nucleotides) was concatenated with any unscaMolded contigs from the error-corrected mabs assembly. These five scaMold-level assemblies have been deposited at DDBJ/ENA/GenBank (accessions TBD).

A scaMold-level genome was identically generated for Mzebra_GT3a. These scaMolds were further aligned to the corrected UMD2a genome (see below) and anchored to linkage groups using D-GENIES59 with the Minimap2 v2.28 aligner and “Many repeats” options. An anchored genome was output using the Query assembled as reference export option.

Note that the mitochondrial DNA sequence in Mzebra_GT3a is the same as the mitochondria assembled in UMD2a. This Whole Genome Shotgun project has been deposited at DDBJ/ENA/GenBank under the accession JBEVYI000000000. The version described in this paper is version JBEVYI010000000.

Correcting errors in M_zebra_UMD2a

We generated Bionano molecules for 3 male and 1 female Metriaclima zebra, assembled Bionano maps, and aligned them to the UMD2a reference genome. We reordered or reoriented contigs that were revealed as misassemblies or inversions, respectively, in all four Bionano maps and corresponded to breakpoints between contigs in UMD2a. By filtering interchromosomal translocation calls in the Bionano Access software, we were able to insert unmapped contigs into the established linkage groups.

Illumina extractions and sequencing

DNA was extracted from fresh or frozen fin tissues for short-read sequencing using the Qiagen MagAttract HMW DNA Kit (Cat. No. 67563) or the Qiagen DNeasy Blood & Tissue Kit (Cat No. 69504) using the Manual Purification of High-Molecular Weight Genomic DNA from Fresh or Frozen Tissue and Purification of Total DNA from Animal Tissues (Spin-Column) protocols respectively. DNA was delivered to the Georgia Tech Molecular Evolution Core where libraries were prepared with the NEBNext® Ultra™ II FS DNA Library Prep Kit for Illumina (NEB #E6177) using the Protocol for FS DNA Library Prep Kit (E7805, E6177) with Inputs ≥100 ng. A small subset of samples was sent to an external collaborator by the Molecular Evolution Core where library preparation was performed using the KAPA HyperPrep Kit (Roche, Material Number: 07962363001). Samples across all runs were sequenced on a NovaSeq 6000 instrument using v1.5 chemistry.

Variant calling and PCA analysis

We provide a link to a github repository (https://github.com/ptmcgrat/CichlidSRSequencing/tree/Kumar_eLife) containing the scripts used for the major analysis in the paper. Because our data is behind a secure Dropbox account, readers will be limited to seeing the exact programs, filters, and parameters used for this manuscript embedded within each script.

Fastq files from Illumina sequencing were converted to UBAM format using the gatk60 FastqToSam algorithm. UBAM files were used for alignment to Mzebra_GT3 using bwa61 mem; the -M and -p flags were used. To carry forward metadata from the unaligned BAMs, gatk MergeBamAlignment was called on the alignments. BAM files were converted to GVCF format using gatk HaplotypeCaller. Variant calling was performed using the GVCF files for the analysis cohort. The gatk GenomicsDBImport and gatk GenotypeGVCFs algorithms were used to generate a master vcf file which was subsequently filtered with gatk VariantFiltration (parameters below). The variants that passed filtering were used for PCA analysis.

PCA was performed with plink262. To avoid overrepresenting species from the cohort, a core subset of the samples was used for PC 1 and 2 calculations (Table S2, column S). Eigenvectors from the whole cohort were then plotted on this PC space and visualized using plotly.

Note that GATK ≥ v4.3.0.0 and python ≥ 3.7 were used for these analyses. Custom python scripts were written to automate and parallelize processing of the samples in the cohorts.

The commands used for variant calling and PCA are the following:

UBAM Generation

Alignment to GT3

Merge metadata from UBAM to BAM

Generate GVCFS

> gatk HaplotypeCaller -R <GT3_FASTA> -I <BAM_FILE> -ERC GVCF -O <GVCF_FILE> -ERC GVCF is used to generate GVCF files that can be used for subsequent joint genotyping

Read Depth Information

Generate GenomicsDB

Joint Genotyping

Note that an average measure of heterozygosity, 0.00175 was used from what was reported in Malinksy et al.30.

Filtering Variants

Variants were filtered by allele frequency, inbreeding coefficient, quality by depth, depth, fisher’s exact test for strand bias, mapping quality, and by excess missingness according to methods published in Malinksky et al.30. We filtered by depth to include variants with a mean depth per sample between 10% and 95% of the distribution of all variant depths.

PCA

For PCA, first a subset of the pass_VCF_file containing the samples we used to create the principal components was generated using bcftools63. Next, the relevant pfiles were generated for both the pass_VCF_file and the subset_samples_VCF. Linkage pruning was only performed for the whole genome and whole chromosome PCAs using the –indep-pairwise 50 5 0.1 flag and parameters. Since inversions inherently link together variants, linkage pruning was not used when restricting the analysis to within inverted regions. A linear scoring system is generated using the subset_samples_VCF pfiles. This scoring system is applied to the genotype matrix for the whole cohort’s samples to scale each sample consistently. The resulting first two eigenvectors per sample from the .sscore file are plotted with plotly.

LD Pruning for Whole genome and whole chromosome PCAs:

No LD Pruning for inverted region PCAs:

Genome alignment

Genomes were aligned using minimap264 (v2.22) to align the query genome to the M_zebra_GT3 reference using default settings. Custom scripts were used to convert the output paf files to dataframe objects, filtering removed alignments that were 1) secondary matches, 2) had less than 30% percent identify, and 3) were less than 10,000 base pairs in length. Further, we excluded any alignments from a contig containing less than 2,000,000 total bp of alignments, to account for repetitive DNA. Alignments were plotted using seaborn. We estimated the intervals containing the breakpoints for the inversions using genome alignments of Aulonocara sp. ‘chitande type north’ Nkhata Bay and M_zebra_GT3 (Table S4).

Phylogenies

To create phylogenies, bcftools was used to filter out non SNV variants from the master vcf file. Bcftools was also used to create individual vcf files for each inversion, filtering out SNVs that fell outside of the inverted region. Vcf2philip65 was used to convert the to the phylip format, using the -m 50 option to filter out SNVs with less than 50 genotyped samples. Iqtree266 was used to create trees and estimate confidence values for each node using the following options: -nt’,‘24’,‘-mem’,‘96G’,‘- v’,‘--seqtype’, ‘DNA’,‘-m’,‘GTR+I+G’,‘-B’,‘1000’. Trees were visualized with iTol67 or ete368.

Nucleotide diversity analysis

To analyze the Pixy between individuals of diMerent species, we used scikit-allel (v1.3.8)69 to calculate the genetic diMerence between each sample using the allel.pairwise_distance function and the citblock for the metric. We restricted the analysis to benthic animals that carried the inversion and compared these to animals of various ecogroups. We used the seaborn function hist plot to create density histograms for each of the inversions.

Fst & pedigree

A vcf file was created for the Aulonocara samples listed in Table S2 (Column I). The PCA approach (described above) was used to genotype each of the samples for the inversion on 10. For the FST analysis, we used scikit-allel (v1.3.8) to perform allel.average_weir_cockerham_fst with a window size of 100. For the heterozygosity analysis, we used the allel.heterozygosity_observed function on both the male and female population, subtracting the average heterozygosity from the male oMspring from the female oMspring.